A Mountaintop View Requires Minimal Sorting: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}

### 2014-11-03 paint-contour-tree.tex
small changes

It is convenient to think of $f$ as a $(d+1)$-dimensional manifold, or alternately consider
$\MM$ in $\RR^{d+1}$ with the last coordinate (i.e. height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
with $f$ gives a set of connected components, the \emph{contours} at height $h$. As the sweeping 
proceeds various events occur: new contours are created or destroyed, contours merge into each other or
split into new components, contours acquire or lose handles. These are the relevant events where 
the topology of the level set changes. The contour tree is a concise representation of all these events.

@@

\Fig{relative} and \Fig{sorting} show examples of simplicial complexes, with heights and their contour trees.
Think of the contour tree edges as pointing downwards, so we can talk of the ``up-degree" and ``down-degree" of vertices.
Leaves of down-degree $1$ are maxima, because a new contour is created at such points. Leaves of up-degree
$1$ are minima, because a contour is destroyed there. Up-degree $2$ vertices are ``joins", where contours
merge, and down-degree $2$ vertices are ``splits".  When $d=2$ (as in \Fig{relative}), contours are closed loops.
In dimensions $d \geq 3$, we may have internal
vertices of both up and down-degree $1$. They correspond to contours (which are $2$-manifolds) gaining or losing handles.

@@

Let us interpret the running time. 
For any path decomposition $P(T)$, obviously $\sum_{p\in P(T)}|p| = |T| = t$, and 
so the running time is $O(t\log t + N)$. But this is tight only when $T$ is tall and thin.  
A better but still obvious upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $P(T)$. 
This is clearly an improvement when the tree is short and fat.
In particular, for any decomposition of a rooted balanced binary tree, this gives $O(t\log\log n + t\alpha(t) + N)$, a significant improvement from $O(t\log t+N)$.
However, our actual bound in this case gives an even further improved $O(t\alpha(t)+N)$.
Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.
(In the example, $t = \Omega(n)$.) The running time promised by \Thm{main-alg}
is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$.

@@

There has been a long series of improved algorithms for computing contour trees.
van Kreveld \etal \cite{kobps-ctsssit-97} presented an $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions. 
Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
The influential paper of Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N)$.
One cannot overstate the importance of this result, since it forms the basis 
of subsequent practical and theoretical work in computational topology applications. 
Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an unconditional $O(N+t\log t)$ algorithm.
The latter result crucially uses monotone paths to improve the running time, and we repeatedly
exploit numerous properties of monotone paths. Carr's thesis shows relationships between monotone paths
in $\MM$ and in its contour tree $\cC(\MM)$.

@@

The $(f(x)+\delta)$-contours in $f|_{B_\eps(x)}$ given by \Def{deg} might actually be the same
contour in $f$. The up-degree (as opposed to \emph{Morse} up-degree) is defined as the number of $(f(x)+\delta)$-contours
that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
When the Morse up-degree is $2$ but the up-degree is $1$, then the level set gains a handle.
When $d=2$, this distinction is not necessary, since any point with $>1$
Morse degree will have some $>1$ degree in $\reeb(\MM)$.

@@

\begin{definition} 
\label{def:int-criticalJoin}
The critical join tree $\jc(\MM)$ is built on the set $V'$ of all critical points. 
The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V'$ in a connected component of $\MM^+_v$
and $v$ is connected to this component (in $\MM$). 
\end{definition}

@@

A major insight of~\cite{csa-cctad-00} is an ingeniously simple linear time procedure to construct
the contour tree from the join and split tree. So the bottleneck is computing these trees. And observe
in \Fig{sorting} that the split vertices form a long path in the join tree (and vice versa). Constructing
these trees forces a global sort of the splits, a redundant computation. Unfortunately (unlike \Fig{sorting}),
the heights of joins and splits may be interleaved in a complex manner. The final merging of~\cite{csa-cctad-00}
to get the contour tree requires having the split vertices in the join tree. Without
this, it is not clear how to get a consistent view of both joins and splits, required for the contour tree.

Our aim is break $\MM$ into smaller pieces, where this redundant computation can be saved.

@@

\begin{figure}[h]\centering
    \includegraphics[width=.18\linewidth]{figs/tent2}%
    \hspace{.9in}
    \includegraphics[width=.18\linewidth]{figs/tent1}
    \caption{On left, downward rain spilling only (each shade of gray represents a different color), producing a grid.  On right, flipping the direction of rain spilling.}
    \label{fig:order}
\end{figure}

\paragraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
endowed with a height to be \emph{minimum dominant} if there only exists a single minima.
In such a complex, there exists a non-ascending path from any point to this unique minima.
(Our real definition is more complicated, and involves simplicial complices
that have many minima.) Analogously, we can define maximum dominant simplicial complices,
and both are collectively called extremum dominant.

We will cut $\MM$ into disjoint extremum dominant pieces, in linear time.
The best way to think of our procedure is a meteorology analogy. Take an arbitrary maxima $x$,
and imagine torrential rain at the maxima. The water flows down, wetting any point that has a non-ascending
from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected,
while there may be numerous disconnected dry parts. The interface between the dry and wet parts
are a set of contours, given the by ``water line". The wet part is clearly maximum dominant, since all
wet points have a non-descending path to $x$. So we can simply cut along the interface
contours to get the wet maximum dominant piece $\MM'$. By our contour surgery theorem,
we are left with a set of disconnected dry parts, and we could recur this procedure on them.

But here's the catch. Every time we cut $\MM$ along a contour, we potentially increase
the complexity of $\MM$. Water flows in the interior of facets, and the interface will naturally
cut some facets. Each cut introduces new vertices, and a poor choice of repeated raining leads
to a large increase in complexity. Consider the left of \Fig{order}. Each raining produces a single
wet and dry piece, and each cut introduces many new vertices. If we wanted to partition this terrain into 
maximum dominant simplicial complices, the final complexity would be forbiddingly large.

@@

\begin{theorem} \label{thm:int-rain} There is an $O(N)$ time procedure that cuts $\MM$ into extremum dominant simplicial complices
$\MM_1, \MM_2, \ldots$. Furthermore, given the set of contour trees $\{\cC(\MM_i)\}$, $\cC(\MM)$ can be constructed in $O(N)$ time.
\end{theorem}

\paragraph{Extremum dominance simplifies contour trees:} We will focus on minimum dominant simplicial complices $\MM$. By \Thm{int-rain}, it suffices
to design an algorithm for contour trees on such inputs. 
For the $d=2$ case, it helps to visualize such an input as a terrain with no minima,
except at a unique boundary face (think of a large boundary triangle that is the boundary).
All the valleys in such a terrain are necessarily joins, and there can be no splits. Look
at \Fig{sorting}. The portion on the left is minimum dominant in exactly this fashion.
Water poured at any critical point must flow to the boundary, and the entire wet portion are connected.
More formally, $\MM^-_v$ is connected for all $v$, so there are no splits.

>>>>

It is convenient to think of $f$ as a $(d+1)$-dimensional manifold, or alternately consider
$\MM$ in $\RR^{d+1}$ with the last coordinate (i.e. height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
with $f$ gives a set of connected components, the \emph{contours} at height $h$. As the sweeping 
proceeds various events occur: new contours are created or destroyed, contours merge into each other or
split into new components, contours acquire or lose handles. More generally, these are the events where 
the topology of the level set changes. The contour tree is a concise representation of all these events.

@@

\Fig{relative} and \Fig{sorting} show examples of simplicial complexes, with heights and their contour trees.
Think of the contour tree edges as pointing downwards, so we can talk of the ``up-degree" and ``down-degree" of vertices.
Leaves of down-degree $1$ are maxima, because a new contour is created at such points. Leaves of up-degree
$1$ are minima, because a contour is destroyed there. Up-degree $2$ vertices are ``joins", where contours
merge, and down-degree $2$ vertices are ``splits".  When $d=2$ (as in \Fig{relative}), contours are closed loops.
In dimensions $d \geq 3$, we may have internal
vertices of both up and down-degree $1$. They correspond to contours (which are $(d-1)$-manifolds) gaining or losing handles.

@@

Let us interpret the running time. 
For any path decomposition $P(T)$, obviously $\sum_{p\in P(T)}|p| = |T| = t$, and 
so the running time is $O(t\log t + N)$. But this is tight only when $T$ is tall and thin.  
A better but still obvious upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $T$. 
This is clearly an improvement when the tree is short and fat.
In particular, for any decomposition of a rooted balanced binary tree, this gives $O(t\log\log n + t\alpha(t) + N)$, a significant improvement from $O(t\log t+N)$.
However, our actual bound in this case gives an even further improved $O(t\alpha(t)+N)$.
Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.
(In the example, $t = \Omega(n)$.) The running time promised by \Thm{main-alg}
is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$.

@@

There has been a long series of improved algorithms for computing contour trees.
van Kreveld \etal \cite{kobps-ctsssit-97} presented an $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions. 
Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
The influential paper of Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
One cannot overstate the importance of this result, since it forms the basis 
of subsequent practical and theoretical work in computational topology applications. 
Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an unconditional $O(N+t\log t)$ algorithm.
The latter result crucially uses monotone paths to improve the running time, and we repeatedly
exploit numerous properties of monotone paths. Carr's thesis shows relationships between monotone paths
in $\MM$ and in its contour tree $\cC(\MM)$.

@@

The $(f(x)+\delta)$-contours in $f|_{B_\eps(x)}$ given by \Def{deg} might actually be the same
contour in $f$. The up-degree (as opposed to \emph{Morse} up-degree) is defined as the number of $(f(x)+\delta)$-contours
that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
When the Morse up-degree is $2$ but the up-degree is $1$, the topology of the level set changes but not by the number of connected components changing.
For example, when $d=3$ this is equivalent to the contour gaining a handle.
When $d=2$, this distinction is not necessary, since any point with $>1$
Morse degree will have some $>1$ degree in $\reeb(\MM)$.

@@

\begin{definition} 
\label{def:int-criticalJoin}
The critical join tree $\jc(\MM)$ is built on the set $\cV$ of all critical points. 
The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $\cV$ in a connected component of $\MM^+_v$
and $v$ is connected to this component (in $\MM$). 
\end{definition}

@@

A major insight of~\cite{csa-cctad-00} is an ingeniously simple linear time procedure to construct
the contour tree from the join and split trees. So the bottleneck is computing these trees. Observe
in \Fig{sorting} that the split vertices form a long path in the join tree (and vice versa). Therefore, constructing
these trees forces a global sort of the splits, an unnecessary computation for the contour tree. Unfortunately, in general (i.e. unlike \Fig{sorting}) 
the heights of joins and splits may be interleaved in a complex manner, and hence the final merging of~\cite{csa-cctad-00}
to get the contour tree requires having the split vertices in the join tree. Without
this, it is not clear how to get a consistent view of both joins and splits, required for the contour tree.

Our aim is to break $\MM$ into smaller pieces, where this unnecessary computation can be avoided.

@@

\begin{figure}[h]\centering
    \includegraphics[width=.18\linewidth]{figs/tent2}%
    \hspace{.9in}
    \includegraphics[width=.18\linewidth]{figs/tent1}
    \caption{On left, downward rain spilling only (each shade of gray represents a different color), producing a grid.  
    Note we are assuming raining was done in reverse sorted order of maxima.  
    On right, flipping the direction of rain spilling.}
    \label{fig:order}
\end{figure}

\paragraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
endowed with a height to be \emph{minimum dominant} if there only exists a single minima.
(Our real definition is more complicated, and involves simplicial complexes
that allow additional ``trivial'' minima.)
In such a complex, there exists a non-ascending path from any point to this unique minima.
Analogously, we can define maximum dominant simplicial complexes,
and both are collectively called extremum dominant.

We will cut $\MM$ into disjoint extremum dominant pieces, in linear time.
The best way to think of our procedure is a meteorology analogy. Take an arbitrary maxima $x$,
and imagine torrential rain at the maxima. The water flows down, wetting any point that has a non-ascending path 
from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected,
while there may be numerous disconnected dry parts. The interface between the dry and wet parts
is a set of contours, given by the ``water line". The wet part is clearly maximum dominant, since all
wet points have a non-descending path to $x$. So we can simply cut along the interface
contours to get the wet maximum dominant piece $\MM'$. By our contour surgery theorem,
we are left with a set of disconnected dry parts, and we can recur this procedure on them.

But here's the catch. Every time we cut $\MM$ along a contour, we potentially increase
the complexity of $\MM$. Water flows in the interior of facets, and the interface will naturally
cut some facets. Each cut introduces new vertices, and a poor choice of repeated raining leads
to a large increase in complexity. Consider the left of \Fig{order}. Each raining produces a single
wet and dry piece, and each cut introduces many new vertices. If we wanted to partition this terrain into 
maximum dominant simplicial complexes, the final complexity would be forbiddingly large.

@@

\begin{theorem} \label{thm:int-rain} There is an $O(N)$ time procedure that cuts $\MM$ into extremum dominant simplicial complexes
$\MM_1, \MM_2, \ldots$. Furthermore, given the set of contour trees $\{\cC(\MM_i)\}$, $\cC(\MM)$ can be constructed in $O(N)$ time.
\end{theorem}

\paragraph{Extremum dominance simplifies contour trees:} We will focus on minimum dominant simplicial complexes $\MM$. By \Thm{int-rain}, it suffices
to design an algorithm for contour trees on such inputs. 
For the $d=2$ case, it helps to visualize such an input as a terrain with no minima,
except at a unique boundary face (think of a large boundary triangle that is the boundary).
All the valleys in such a terrain are necessarily joins, and there can be no splits. Look
at \Fig{sorting}. The portion on the left is minimum dominant in exactly this fashion.
Water poured at any critical point must flow to the boundary, and the entire wet portion is connected.
More formally, $\MM^-_v$ is connected for all $v$, so there are no splits.
